Semi-quantitative petrographic characterization of Bronze Age quartzite tools – a multivariate approach

### 2022-01-20 analysis/paper/paper.Rmd
Updated code and text

#Import xrf data, set empty fields to NA
XRF.csv <-
  read.csv2("./analysis/data/raw_data/XRF/XRF_quantitative.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))

#Import nir data, set empty fields to NA
nir.csv <-
  read.csv2("./analysis/data/derived_data/NIR_master_averaged_20210914.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))

#Import raman data, set empty fields to NA
raman.csv <-
  read.csv2("./analysis/data/derived_data/raman_baseline_corrected_20210826.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))

#Import sample list, set empty fields to NA
samples <-
  read.csv2("./analysis/data/raw_data/bifacial_points.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na.strings=c("", "NA"))

@@

In regions where local flint and obsidian sources are non-existent, such as Northern Fennoscandia, other materials have been the primary target for lithic production^[This is not to say that flint was not available or used in these regions] [@baudou78; @rankama_etal06; @olausson_etal12]. The last decades have seen an increase in interest for non-flint lithics and raw material studies in  European archaeology, providing much needed development in a previously neglected field [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19]. With the lack of any natural geological source of flint in Northern Sweden both quartz and quartzite are instrumental for the inference of settlement and mobility within the region. While quartz and quartzite technology have been part of a number of studies over the years, there have been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10]. One reason for the lack of methodological development in North Swedish research could be the complex geological background of the region and the abundance of moraine formations [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16]. The degree to which geological materials have been mixed and transported demands considerable effort in order to reconstruct the events that have formed the modern day landscape. While tentaive interpretations of lithic material sourcing tends to describe strategies leaning towards direct procurement from the mountain zone [@forsberg85; -@forsberg12], both quartz and quartzite could potentially have been sourced locally from exposed veins in the bedrock, or as loose rocks and pebbles in the numerous moraine formations, as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06 pp.249-250]. Due to this ambiguous nature of availability of raw materials provenience studies are complex and fraught with uncertainty.

@@

##Aims

In the current study a dataset of quartz and quartzite artefacts from museum collections in Västerbotten, Sweden, have been analysed using the three above mentioned spectroscopic instrumentations (NIRS, Raman, XRF). The aim of the study is to evaluate their viability for fingerprinting quartz and quartzite material in the North Swedish region, where there currently is a lack of reference material as well as a complex geological setting. The study is a part of a larger project wherein models of settlement and mobility are being evaluated and developed in regards to raw material procurement and use. As flint is a non-local material to Northern Sweden it follows that any find of this material within the region most likely have been brought there through anthropogenic activity. In contrast, determining the degree to which the distribution pattern of quartz and quartzite have been influenced by cultural and/or environmental factors however, is far more difficult. While there is a good understanding of the geological makeup of the Scandes [@stephens_bergmanweihed20], the quartzite itself remains largely uncharacterized, with few reference samples available for comparison. The glacial impact on the landscape further compounds the issue as geological material has been transported and deposited by glacial movements, as well as subsequent postglacial processes. Although models for settlement and mobility can be used to test ideas of material distribution as influenced by cultural processes [e.g. @forsberg85], it is also necessary to attempt to estimate the degree of environmental impact on this pattern.

This will be achieved via Principal Component Analysis (PCA), where the observed groupings of the material in the model will be related to their spectral features.

@@

#Reorder factor levels
Points2$type <-
  factor(Points2$type, levels=c("Preform", "Point fragment", "Point"))
```

@@

The current study is based on spectroscopic surface measurements collected on quartz and quartzite points recovered from sites within the County of Västerbotten, Sweden. A dataset of `r length(sample_uniq)` points, point fragments and preforms made from quartz and quartzite material stored at the collections at Västerbotten (www.vbm.se) and Skellefteå (www.skellefteamuseum.se) museums were analysed using NIRS, raman and XRF spectroscopy. The material features a mainly black, grey or white hue, as well as a subset of semi-translucent material with a beige tint to almost fully translucent. The colour of the material has been classified both according to the Munsell system, as well as sorted into general categories of hue for visualization purposes: *dark* (~ N1-N4), *light* (~ N5-N8), *white* (~ N9), *colourless* (lack of colour, opaque to translucent). A number of these points have, according to prior classifications, been worked using bifacial technology. While there currently is no local chronology available for this artefact type in the North Swedish region^[For comparison, Mjæreum -@mjaerum12 has put forth a chronology for bifacial points in southwestern Norway], there have been attempts[@baudou92 p.99; @forsberg85 p.5; @forsberg89]. Taken together the research suggests a period of use that falls between ca. 3 800 - 2 000 BP [@kaeck09 p.60]. Similarly, studies of raw material use at sites dating to this period seems to indicate an increase in the use of quartzite [@baudou78].

@@

The XRF data was log-transformed, mean-centered and scaled to unit variance before PCA modelling. As the resulting dataset also features values below the Limit of Detection (LOD), these were treated as NULL (or "missing values"), during PCA. Since missing values are not permitted during PCA these had to imputed before the analysis, which was performed using the *missMDA* R package [@josse_husson16] developed for datasets with incomplete data. Two methods for imputation can be implemented, iterative or regularized, and since the former suffers when there is a large number of missing values the latter was used.

@@

```{r vis-nir-pca, fig.width=20, fig.height=12, fig.cap="PCA score plots for the 400 nm – 750 nm band of the NIR data. Top: PC 1 and PC 2, bottom: PC 1 and PC 3 PC 1 seems to group dark-light material on opposite sides of the plots while PC 2 does little to change the strucure. PC 3 seems to position material with a blue-gray tint towards the top of the score plot."}

@@

@@

The resulting PCA model (fig. \@ref(fig:raman-pca)) shows a clear trend along PC 1 (81.5 %) related to the quartz and D - G peaks. PC 2 and PC 3, however, mainly describes the remaining noise left in the spectra after detrending and preprocessing the data. Interestingly, the loading for PC 4 does include the Al-O-Si peaks around the 700 cm^-1^ - 800 cm^-1^ band, which helps in separating out those samples where this feature is present. The measurements with mainly quartz peaks gravitates towards the negative side of the PC 1, and those where the graphite peaks are dominant towards the opposite side. It is also clear that the darker material feature more pronounced graphite peaks, whereas the lighter and more colourless objects mainly feature quartz peaks. As in the case of the NIR data, however, there are exceptions that break this trend, as well as a “muddled” middle group with both well-pronounced graphite and quartz peaks. When PC 4 is added the samples with quartz and Al-O-Si peaks gravitate towards the bottom left, separating themselves from the "quarts" group.

>>>>

#Import xrf data, set empty fields to NA
XRF.csv <-
  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/raw_data/XRF/XRF_quantitative.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))

#Import nir data, set empty fields to NA
nir.csv <-
  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/derived_data/NIR_master_averaged_20220119.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))

#Import raman data, set empty fields to NA
raman.csv <-
  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/derived_data/raman_baseline_corrected_20220119.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))

#Import sample list, set empty fields to NA
samples <-
  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/raw_data/bifacial_points.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na.strings=c("", "NA"))

@@

The last decades have seen an increase in the interest for non-flint lithics and raw material studies in European archaeology, providing much needed development in a previously neglected field [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19]. In regions where local flint and obsidian sources are non-existent, such as Northern Fennoscandia, other materials have been the primary target for lithic production^[This is not to say that flint was not available or used in these regions] (e.g. quartz, quartzite, slate) [@baudou78; @rankama_etal06; @olausson_etal12]. With the lack of any natural geological source of flint in Northern Sweden both quartz and quartzite are instrumental for the inference of settlement and mobility within the region. While quartz and quartzite technology have been part of a number of studies over the years, there have been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10]. One reason for the lack of methodological development in North Swedish research could be the complex geological background of the region and the abundance of moraine formations [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16]. The degree to which geological materials have been mixed and transported demands considerable effort in order to reconstruct the events that have formed the modern day landscape. While tentaive interpretations of lithic material sourcing tends to describe strategies leaning towards direct procurement from the mountain zone [@forsberg85; -@forsberg12], both quartz and quartzite could potentially have been sourced locally from exposed veins in the bedrock, or as loose rocks and pebbles in the numerous moraine formations, as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06 pp.249-250]. Due to this ambiguous nature of availability of raw materials provenience studies are complex and fraught with uncertainty.

@@

## Aims

In the current study a dataset of quartz and quartzite artefacts from museum collections in Västerbotten, Sweden, have been analysed using the three above mentioned spectroscopic instrumentations (NIRS, Raman and XRF). The aim of the study is to evaluate their viability for fingerprinting quartz and quartzite material in the North Swedish region based on surface measurements. The study is a part of a larger project wherein models of settlement and mobility are being evaluated and developed in regards to raw material procurement and use. As flint is a non-local material to Northern Sweden it follows that any find of this material within the region most likely have been brought there through anthropogenic activity. In contrast, determining the degree to which cultural and/or environmental factors have influenced the distribution pattern of quartz and quartzite artefacts, however, is far more difficult. While there is a good understanding of the geological makeup of the Scandes [@stephens_bergmanweihed20], the quartzite itself remains largely uncharacterized, with few reference samples available for spectral comparison. The glacial impact on the landscape further compounds the issue as geological material has been transported and deposited by glacial movements, as well as subsequent postglacial processes. Although models for settlement and mobility can be used to test ideas of material distribution as influenced by cultural processes [e.g. @forsberg85], it is also necessary to attempt to estimate the degree of environmental impact on this pattern.

In order to engage in any meaningful analysis, however, it is first necessary to distinguish between different quartz and quartzite groups. Principal Component Analysis (PCA) has been used to model the spectral data in order to discern material groupings related to different spectral features and their contributions to the Principal Components (PC). The results form the basis of a discussion of the viability of using non-destructive spectroscopy in petrographic characterization of quartz and quartzite material, and thus future efforts of developing models for raw material sourcing in Northern Sweden.

@@

#Reorder type factor levels
Points2$type <-
  factor(Points2$type, levels=c("Preform", "Point fragment", "Point"))

#Reorder parish factor levels
Points2$parish <- 
  factor(Points2$parish, levels = c("Bjurholm","Bureå","Burträsk", "Bygdeå", "Degerfors", "Dorotea", "Lycksele", "Norsjö", "Skellefteå", "Sorsele", "Stensele", "Tärna", "Umeå", "Vännäs", "Vilhelmina", "Åsele"))
```

@@

The current study is based on spectroscopic surface measurements collected on quartz and quartzite points recovered from sites within the County of Västerbotten, Sweden. A dataset of `r length(sample_uniq)` points, point fragments and preforms made from quartz and quartzite material stored at the collections at Västerbotten (www.vbm.se) and Skellefteå (www.skellefteamuseum.se) museums were analysed using NIRS, raman and XRF spectroscopy. The material features a mainly black, grey or white hue, as well as a subset of semi-translucent material with a beige tint to almost fully translucent. The colour of the material has been classified both according to the Munsell system, as well as sorted into general categories of hue for visualization purposes: *dark* (~ N1-N4), *light* (~ N5-N8), *white* (~ N9), *colourless* (lack of colour, opaque to translucent). A number of these points have, according to prior classifications, been worked using bifacial technology. While there currently is no local chronology available for this artefact type in the North Swedish region^[For comparison, Mjæreum -@mjaerum12 has put forth a chronology for bifacial points in southwestern Norway], there have been attempts [@baudou92 p.99; @forsberg85 p.5; @forsberg89]. Previous research suggests a period of use that falls between ca. 3 800 - 2 000 BP [@kaeck09 p.60]. Similarly, studies of raw material use at sites dating to this period seems to indicate an increase in the use of quartzite [@baudou78].

@@

The XRF data was log-transformed, mean-centered and scaled to unit variance before PCA modelling. As the resulting dataset also features values below the Limit of Detection (LOD), these were treated as NULL (or "missing values"), during PCA. Since missing values are not permitted during PCA these had to be imputed before the analysis, which was performed using the *missMDA* R package [@josse_husson16] developed for datasets with incomplete data. Two methods for imputation can be implemented, iterative or regularized, and since the former suffers when there is a large number of missing values the latter was used.

\pagebreak

@@

@@

```{r vis-nir-pca, fig.width=20, fig.height=12, fig.cap="PCA score plots for the 400 nm – 750 nm band of the NIR data. Top: PC 1 and PC 2, bottom: PC 1 and PC 3. PC 1 seems to group dark-light material on opposite sides of the plots while PC 2 does little to change the strucure. PC 3 seems to position material with a blue-gray tint towards the top of the score plot."}

@@

The resulting PCA model (fig. \@ref(fig:raman-pca)) shows a clear trend along PC 1 (81.5 %) related to the quartz and D - G peaks. PC 2 and PC 3, however, mainly describes the remaining noise left in the spectra after detrending and preprocessing the data. Interestingly, the loading for PC 4 does include the Al-O-Si peaks around the 700 cm^-1^ - 800 cm^-1^ band, which helps in separating out those samples where this feature is present. The measurements with mainly quartz peaks gravitates towards the negative side of PC 1, and those where the graphite peaks are dominant towards the opposite side. It is also clear that the darker material feature more pronounced graphite peaks, whereas the lighter and more colourless objects mainly feature quartz peaks. As in the case of the NIR data, however, there are exceptions that break this trend, as well as a “muddled” middle group with both well-pronounced graphite and quartz peaks. When PC 4 is added the samples with quartz and Al-O-Si peaks gravitate towards the bottom left, separating themselves from the "quarts" group.
